Outline of Ideas for BTC Transaction Frequency

### 2015-10-02 outline/outline.tex
Sketched a class of models including measurement error on arrival times and block size

\section{The Basic Problem}
Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $t$ index time in minutes, and $N(t)$ denote the number of blocks discovered by time $t$. Then for $t>s$, the number of blocks discovered between time $s$ and time $t$ is Poisson distributed, that is $N(t) - N(s) \sim Poi((t-s)/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\tau_k - \tau_{k-1} \sim exp(1/10)$. This part of the problem is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work. Suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_t$ so that the total amount of data between time $s$ and time $t$, $t>s$, is the stochastic integral $x_t - x_s = \int_s^tdx_t$. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k+1}}dx_t$. The quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k+1}}dx_t > 1\right)
\]
for some choice of $k$. If we discretize the processes so that $t=1,2,\dots$ denotes the minute, this becomes
\[
P\left(\sum_{t=\tau_{k}}^{\tau_{k+1}}dx_t > 1\right).
\]
For some distributional forms for the process $dx_t$ it will be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\sum_{t=\tau_{k}^{(i)}}^{\tau_{k+1}^{(i)}}dx_{t}^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function.

The only piece we need to do this is the distribution of the $dx_t$'s. We have two major options here --- we can choose prespecified distributions in order to see how different distributional assumptions affect these probabilities, or we can attempt to estimate the distribution of the transaction data. The former is easy and worth doing just to gather some idea of how different forms of uncertainty affect these probabilities. In the easiest case, we can assume that the $dx_t$'s are fixed at a mean rate of data arrival and analytically compute the probability that a block fills up, showing how this changes depending on the mean data rate. In the most complicated case, we can specify a dynamic model which includes seasonality and other forms of temporal dependence with, again, some fixed mean rate.

Ideally, we would like to fit this sort of model to the BTC transaction data. Doing so will give us more a more empirically plausible distribution with which to compute the probability that a block fills up. Depending on the model, we may be able to predict the growth of BTC transaction data as a side effect, and this may be interesting in its own right.

>>>>

\section{The Basic Model}
Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} exp(1/10)$ for $k=1,2,\dots,K$. This part of the model is determined for us, but determining the how the number of transactions is distributed or, more precisely, how the amount of transaction {\it data} is distributed requires more work.

Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the upper bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Let $\tau_k$ denote the actual time of the arriaval of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed time in the blockchain. Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.

Most of the interesting modeling choices come from modeling $dx_\tau$. Conventional Brownian motion is inappropriate since the increments should be nonnegative. On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.  For example, the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error since this simplifies Markov chain Monte Carlo methods that may be used to fit the model. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, but most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} dsitribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.

For example, consider the following parameterized modeling using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\alpha\delta_k,\beta)\\
\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

\section{Put this where it needs to go}

The quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
\]
for some choice of $k$.

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. This formula assumes that we can perfectly simulate the continous sample paths of $dx_\tau$, which may require further approximation.
